Storm surge modelling in the Strait of Georgia

### 2014-10-20 RoughWork/manuscript.tex
Expanded on discussion section. Added other authors

\title{Storm surge modelling in the Strait of Georgia}
\author{Nancy Soontiens, Susan Allen, Doug Latornell, Kate Le Sou\"{e}f, Idalia Machuca}
\date{\today}

@@

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step is and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restrictive step through the vertical CFL condition.

@@

@@

Hindcasts of significant events in February 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude (>50 \%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up local differences in the surge amplitude, resulting in local variations in the spatial distribution of the surge. Generally, surge amplitudes in the Strait of Georgia are higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to tide-surge interaction were most important after the San Juan and Gulf Islands.

Several factors considered in other storm surge modelling studies have been neglected. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season.

Since the Strait of Georgia is a large, semi-enclosed body of water storm surges can behave and evolve differently than in other coastal regions. To be continued...

@@

\begin{table}[h]
\centering 
\begin{tabular}{l l l l c c c c c c} 
\hline 
& & & & \multicolumn{3}{c}{M$_2$} & \multicolumn{3}{c}{K$_1$} \\ 
\hline 
            & Station name     & Longitude & Latitude & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) \\
\hline 
1 & Port Renfrew & -124.43 & 48.55 & 1.16 & -9.4 & 16.63  & 1.03 & -4.94 & 4.17 \\
2 & Sheringham Point & -123.92 & 48.38 & 1.19 & -16.47 & 17.95  & 1 & -4.44 & 4.24 \\
3 & Pedder Bay & -123.55 & 48.33 & 0.96 & -27.39 & 15.91  & 0.99 & -2.41 & 2.66 \\
4 & Esquimalt & -123.43 & 48.43 & 0.91 & -23.09 & 14.40  & 0.99 & -1.04 & 1.21 \\
5 & Victoria & -123.37 & 48.42 & 0.89 & -17.77 & 11.62  & 1.04 & -1.83 & 3.22 \\
6 & Clover Point & -123.35 & 48.41 & 0.83 & -17.95 & 13.39  & 1.03 & -2.23 & 3.14 \\
7 & Finnerty Cove & -123.3 & 48.47 & 0.83 & -8.24 & 9.47  & 1.02 & -2.19 & 3.25 \\
8 & Fulford Harbour & -123.45 & 48.77 & 0.82 & -3.08 & 10.78  & 1 & -0.63 & 0.83 \\
9 & Tumbo Channel & -123.11 & 48.79 & 1.05 & 2.03 & 4.43  & 1.04 & 0.66 & 3.34 \\
10 & Patos Island & -122.97 & 48.78 & 1.03 & 0.26 & 2.32  & 1.04 & -1.5 & 3.92 \\
11 & Whaler Bay & -123.33 & 48.89 & 1 & -0.76 & 1.10  & 1.01 & -0.48 & 1.19 \\
12 & Tsawwassen & -123.13 & 48.99 & 1 & 0.92 & 1.31  & 1.01 & 0.19 & 1.09 \\
13 & Sandheads & -123.3 & 49.1 & 1.02 & -1.05 & 2.58  & 1.04 & -0.59 & 3.15 \\
14 & Point Grey & -123.27 & 49.25 & 0.98 & -3.31 & 5.83  & 0.96 & -0.68 & 3.63 \\
15 & Point Atkinson & -123.25 & 49.33 & 1.01 & -0.48 & 1.07  & 1 & -0.56 & 0.94 \\
16 & Gibsons Landing & -123.5 & 49.4 & 0.98 & 0.75 & 2.34  & 1 & 0.96 & 1.47 \\
17 & Winchelsea & -124.08 & 49.3 & 1.01 & -0.19 & 0.78  & 1.01 & 0.51 & 1.30 \\
18 & Halfmoon Bay & -123.92 & 49.52 & 1.01 & -0.15 & 0.83  & 1.01 & 0.15 & 0.72\\ 
19 & Irvines Landing & -124.05 & 49.63 & 1 & -0.32 & 0.59  & 1.01 & 0.24 & 1.12 \\
20 & Powell River & -124.55 & 49.86 & 1 & -1.72 & 3.05  & 1 & -0.01 & 0.43 \\
21 & Little River & -124.92 & 49.74 & 1.03 & 0.07 & 2.54  & 1.01 & 0.33 & 0.73\\ 
22 & Lund & -124.76 & 49.98 & 1.02 & -2.52 & 4.98  & 1.02 & -0.95 & 2.45 \\
23 & Twin Islets & -124.93 & 50.03 & 1.02 & -2.05 & 4.22  & 1.01 & -0.24 & 0.72\\ 
24 & Campbell River & -125.24 & 50.04 & 0.99 & 0.51 & 1.12  & 1 & 0.94 & 1.39 \\
25 & Maude Island E & -125.33 & 50.13 & 1.13 & -10.02 & 12.62  & 0.97 & -1.44 & 2.89\\ 
26 & Nymphe Cove & -125.36 & 50.13 & 1.02 & -5.43 & 6.04  & 0.97 & -0.29 & 2.02 \\
27 & Seymour Narrows & -125.34 & 50.13 & 0.63 & 31.31 & 53.40  & 1.14 & 10.91 & 17.28 \\
28 & Brown Bay & -125.37 & 50.16 & 0.98 & -5.63 & 9.26  & 0.98 & -0.04 & 1.17 \\
29 & Chatham Point & -125.44 & 50.33 & 1.09 & -5.49 & 12.18  & 0.98 & -3.18 & 3.85 \\
30 & Kelsey Bay & -125.96 & 50.39 & 1 & -1.69 & 3.50  & 0.98 & -0.39 & 1.00 \\
31 & Yorke Island & -125.97 & 50.44 & 0.99 & 1.59 & 3.32  & 1.01 & 1.23 & 1.37\\ 
\hline 
&   Mean Error (cm/deg) & & & 4.23  & 6.5 & 8.05   & 1.56 & 1.49 & 2.58 \\
&   RMS Error (cm/deg)  & & & 7.67  & 10.73 & 12.73 & 2.41 & 2.57 & 3.91 \\
\hline  
\end{tabular}
\caption{Comparisons of modelled and observed amplitudes and phases. $R$ is the ratio of modelled to observed amplitude. $\Delta \phi$ is the difference between modelled and observed phase. $D$ is the complex difference. }
\label{tab:comparison} 
\end{table}

\section{Acknowledgements}\label{sec:ack}
We would like to thank Michael Foreman, Diane Masson, and Luc Fillion for providing data used in model set up and evaluation. This project is funded by the Marine Environmental Observation Prediction and Response Network (MEOPAR), a Network of Centres of Excellence of Canada.  Others like Keith, Youyu, Vasily, JP, Rich, Mark?

>>>>

\title{Storm surge modelling in the Strait of Georgia}
\author{Nancy Soontiens, Susan Allen, Doug Latornell, Kate Le Sou\"{e}f, Idalia Machuca,
\\ Youyu Lu, Keith Thompson, Jean-Philippe Paquin, Vasily Korabel}
\date{\today}

@@

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generalll, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restrictive time step through the vertical CFL condition.

@@

\subsection{An Additional Hindcast: November 16, 2006}

@@

Hindcasts of significant events in February 2006, November 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude (>50 \%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up spatial differences in the surge amplitude, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are usually higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important after the San Juan and Gulf Islands. The modelled surge amplitude increased by up to 10 \% at Point Atkinson when the tide-surge interaction was neglected.

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

Since the Strait of Georgia is a large, semi-enclosed body of water storm surges can behave and evolve differently than in other coastal regions. For instance, the complicated coastlines and local wind patterns result in important spatial variations in the surge amplitude. As such, strong winds caused by a large storm is not the only factor to be considered in community preparedness for surge conditions. The geometry of the Strait and local wind patterns render communities on the mainland side of the southern Strait of Georgia more susceptible to surges. Additionally, commmunities such as Campbell River are at risk due to the long fetch and predominately southerly winds during the winter months.

Finally, difficulties in reproducing accurate M$_2$ tidal harmonics in both the Strait of Georgia and the Strait of Juan de Fuca have been highlighted in this study among others \citep{stronach1993update, foreman2004m}. Future work will attempt to improve model accuracy by examining mixing parametrizations and energy dissipation due to bottom friction and their effect on tidal haromnics. Additionally, a biological-chemical modelling compenent will be added.

@@

\begin{table}[h]
\centering 
\begin{tabular}{l l l l c c c c c c} 
\hline 
& & & & \multicolumn{3}{c}{M$_2$} & \multicolumn{3}{c}{K$_1$} \\ 
\hline 
            & Station name     & Longitude & Latitude & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) & $R$ & $ \Delta \phi$ (deg) & $D$ (cm) \\
\hline 
1 & Port Renfrew & -124.43 & 48.55 & 1.16 & -9.4 & 16.63  & 1.03 & -4.94 & 4.17 \\
2 & Sheringham Point & -123.92 & 48.38 & 1.19 & -16.47 & 17.95  & 1 & -4.44 & 4.24 \\
3 & Pedder Bay & -123.55 & 48.33 & 0.96 & -27.39 & 15.91  & 0.99 & -2.41 & 2.66 \\
4 & Esquimalt & -123.43 & 48.43 & 0.91 & -23.09 & 14.40  & 0.99 & -1.04 & 1.21 \\
5 & Victoria & -123.37 & 48.42 & 0.89 & -17.77 & 11.62  & 1.04 & -1.83 & 3.22 \\
6 & Clover Point & -123.35 & 48.41 & 0.83 & -17.95 & 13.39  & 1.03 & -2.23 & 3.14 \\
7 & Finnerty Cove & -123.3 & 48.47 & 0.83 & -8.24 & 9.47  & 1.02 & -2.19 & 3.25 \\
8 & Fulford Harbour & -123.45 & 48.77 & 0.82 & -3.08 & 10.78  & 1 & -0.63 & 0.83 \\
9 & Tumbo Channel & -123.11 & 48.79 & 1.05 & 2.03 & 4.43  & 1.04 & 0.66 & 3.34 \\
10 & Patos Island & -122.97 & 48.78 & 1.03 & 0.26 & 2.32  & 1.04 & -1.5 & 3.92 \\
11 & Whaler Bay & -123.33 & 48.89 & 1 & -0.76 & 1.10  & 1.01 & -0.48 & 1.19 \\
12 & Tsawwassen & -123.13 & 48.99 & 1 & 0.92 & 1.31  & 1.01 & 0.19 & 1.09 \\
13 & Sandheads & -123.3 & 49.1 & 1.02 & -1.05 & 2.58  & 1.04 & -0.59 & 3.15 \\
14 & Point Grey & -123.27 & 49.25 & 0.98 & -3.31 & 5.83  & 0.96 & -0.68 & 3.63 \\
15 & Point Atkinson & -123.25 & 49.33 & 1.01 & -0.48 & 1.07  & 1 & -0.56 & 0.94 \\
16 & Gibsons Landing & -123.5 & 49.4 & 0.98 & 0.75 & 2.34  & 1 & 0.96 & 1.47 \\
17 & Winchelsea & -124.08 & 49.3 & 1.01 & -0.19 & 0.78  & 1.01 & 0.51 & 1.30 \\
18 & Halfmoon Bay & -123.92 & 49.52 & 1.01 & -0.15 & 0.83  & 1.01 & 0.15 & 0.72\\ 
19 & Irvines Landing & -124.05 & 49.63 & 1 & -0.32 & 0.59  & 1.01 & 0.24 & 1.12 \\
20 & Powell River & -124.55 & 49.86 & 1 & -1.72 & 3.05  & 1 & -0.01 & 0.43 \\
21 & Little River & -124.92 & 49.74 & 1.03 & 0.07 & 2.54  & 1.01 & 0.33 & 0.73\\ 
22 & Lund & -124.76 & 49.98 & 1.02 & -2.52 & 4.98  & 1.02 & -0.95 & 2.45 \\
23 & Twin Islets & -124.93 & 50.03 & 1.02 & -2.05 & 4.22  & 1.01 & -0.24 & 0.72\\ 
24 & Campbell River & -125.24 & 50.04 & 0.99 & 0.51 & 1.12  & 1 & 0.94 & 1.39 \\
25 & Maude Island E & -125.33 & 50.13 & 1.13 & -10.02 & 12.62  & 0.97 & -1.44 & 2.89\\ 
26 & Nymphe Cove & -125.36 & 50.13 & 1.02 & -5.43 & 6.04  & 0.97 & -0.29 & 2.02 \\
27 & Seymour Narrows & -125.34 & 50.13 & 0.63 & 31.31 & 53.40  & 1.14 & 10.91 & 17.28 \\
28 & Brown Bay & -125.37 & 50.16 & 0.98 & -5.63 & 9.26  & 0.98 & -0.04 & 1.17 \\
29 & Chatham Point & -125.44 & 50.33 & 1.09 & -5.49 & 12.18  & 0.98 & -3.18 & 3.85 \\
30 & Kelsey Bay & -125.96 & 50.39 & 1 & -1.69 & 3.50  & 0.98 & -0.39 & 1.00 \\
31 & Yorke Island & -125.97 & 50.44 & 0.99 & 1.59 & 3.32  & 1.01 & 1.23 & 1.37\\ 
\hline 
&   Mean Error (cm/deg) & & & 4.23  & 6.5 & 8.05   & 1.56 & 1.49 & 2.58 \\
&   RMS Error (cm/deg)  & & & 7.67  & 10.73 & 12.73 & 2.41 & 2.57 & 3.91 \\
\hline  
\end{tabular}
\caption{Comparisons of modelled and observed amplitudes and phases. $R$ is the ratio of modelled to observed amplitude. $\Delta \phi$ is the difference between modelled and observed phase. $D$ is the complex difference. The mean error and RMS error are calculated as the difference between the modelled quantity and observered quantity. }
\label{tab:comparison} 
\end{table}

\section{Acknowledgements}\label{sec:ack}
We would like to thank Michael Foreman, Diane Masson, Luc Fillion Rich Pawlowicz, and Mark Halverson for providing data used in model set up and evaluation. This project is funded by the Marine Environmental Observation Prediction and Response Network (MEOPAR), a Network of Centres of Excellence of Canada.
